Stochastic blockmodeling of relational event dynamics

### 2012-05-31 docs/intro.tex
First few paragraphs in intro.

Statistical methods for analyzing network data are becoming increasingly useful for studying phenomena ranging from online social behavior to protein interactions \cite{Goldenberg2009}.
One unifying idea is that the nodes of these networks are differentiated by rates of interacting with other parts of the network [TODO], where this variation in rates may be associated with some unobserved quantity, e.g. a covariate such as age or sex, or perhaps shared [TODO].
Stochastic blockmodels \cite{Nowicki2001, Kemp, Ishiguro2010, Rodriguez} are a class of statistical models for static network data that employ latent variables to model this heterogeneity by 1) assuming each node in the network belongs to some block (or cluster) and 2) parameterizing the probability of edges between nodes of each block. 
%PS: much of this past work is for static network data. Do we want to make the observation here that we are thinking about blockmodels for *rate* data here in our discussion and in our example in figure 1 - if we don't some readers/reviewers might get a little confused. For example we could say that instead of a binomial model on edges we can think of a Poisson process yielding counts over time, where we can still think about blockmodels. 
%PS: also, we introduce the term "rates" above without really defining it, and then go on to talk about "mean rate" below. So I think we definitely need a sentence somewhere above that makes it clear we are talking about edges that instantaneously appear and disappear over time at some rate, rather than static edges.

If we are interested in studying a network of events, there are other important aspects of the dynamics in addition to the mean rate. 
For example, human conversation often has an increased propensity for reciprocated events.
%PS: wording of next sentence is a little odd - not sure I understand what it means...
Shared tendencies based on other observed or unobserved quantities is of interest (who shares this tendency).
[TODO: Example of a shared dynamic... perhaps sticking with the reciprocity bit.]
Figure \ref{fig:introexample} shows a simple example of how traditional blockmodels are unable to make inferences these types of dynamic phenomena.

>>>>

Statistical methods for analyzing network data are becoming increasingly useful for studying phenomena ranging from online social behavior to protein  interactions \cite{Goldenberg2009}.
We often  observe events occurring between nodes over time, and a common goal is to understand the differences in the total number of interactions --- perhaps with respect to information known about the nodes.
%In this case, we have a count for the total activity of each dyad, and one may analyze it has a static network. 
%Methods for static networks such as a 
Stochastic blockmodels \cite{Nowicki2001, Kemp, Ishiguro2010, Rodriguez} are a class of techniques which can model this heterogeneity by assuming latent subgroups of nodes have similar tendencies in their group-wise rate of interaction.

Network-based event data are powerful, however, because we enable the study of the processes governing these events.
Theories about the dynamics among nodes will be often be difficult to study when only using the marginal rate of events as illustrated in Figure \ref{fig:introexample}
For example, a common aspect of human communication is \emph{reciprocity} --- after person A speaks to person B, B subsequently has a higher propensity to speak to A.  
These sorts of dynamics may be studied with event data, but again, our focus may be on the differences in these dynamics among the nodes in the network.  
In the context of studying interactions among students, one might be interested particular conversational tendencies are associated with covariates about the student or the classroom.  
A standard blockmodel only provides a means for studying the structure of marginal rates of communication, but not the shared tendencies in the dynamics of the sequence of events.

%When such .  One unifying idea is that the nodes of these networks are differentiated by rates of interacting with other parts of the network [TODO], where this variation in rates may be associated with some unobserved quantity, e.g. a covariate such as age or sex, or perhaps shared [TODO].
%Stochastic blockmodels \cite{Nowicki2001, Kemp, Ishiguro2010, Rodriguez} are a class of statistical models for static network data that employ latent variables to model this heterogeneity by .  Such techniques are readily applied to count data, [TODO: These let us study how shared behavior is associated with other covariates.]

%1) assuming each node in the network belongs to some block (or cluster) and 2) parameterizing the probability of a tie between members

%PS: much of this past work is for static network data. Do we want to make the observation here that we are thinking about blockmodels for *rate* data here in our discussion and in our example in figure 1 - if we don't some readers/reviewers might get a little confused. For example we could say that instead of a binomial model on edges we can think of a Poisson process yielding counts over time, where we can still think about blockmodels. 
%PS: also, we introduce the term "rates" above without really defining it, and then go on to talk about "mean rate" below. So I think we definitely need a sentence somewhere above that makes it clear we are talking about edges that instantaneously appear and disappear over time at some rate, rather than static edges.

%If we are interested in studying a network of events, there are other important aspects of the dynamics in addition to the mean rate. 
%For example, human conversation often has an increased propensity for reciprocated events.
%PS: wording of next sentence is a little odd - not sure I understand what it means...
%Shared tendencies based on other observed or unobserved quantities is of interest (who shares this tendency).
%[TODO: Example of a shared dynamic... perhaps sticking with the reciprocity bit.]
%Figure \ref{fig:introexample} shows a simple example of how traditional blockmodels are unable to make inferences these types of dynamic phenomena.
